*SB0358.2*
Reprinted
February 1, 2022
SENATE BILL No. 358
_____
DIGEST OF SB 358 (Updated January 31, 2022 5:26 pm - DI 101)
Citations Affected: IC 24-15.
Synopsis: Consumer data protection. Establishes a new article in the
Indiana Code concerning consumer data protection, to take effect
January 1, 2025. Sets forth the following within the new article: (1)
Definitions of various terms that apply throughout the article. (2)
Exemptionsfromthe bill'srequirements concerningthe responsibilities
of controllers of consumers' personal data. (3) The rights of an Indiana
consumer to do the following: (A) Confirm whether or not a controller
is processing the consumer's personal data.(B)Correct inaccuraciesin
the consumer's personal data that the consumer previously provided to
a controller. (C) Delete the consumer's personal data held by a
controller. (D) Obtain a copy or representative summary of the
consumer's personal data that the consumer previously provided to the
controller. (E) Opt out of the processing of the consumer's personal
data for certain purposes. (4) The responsibilities of controllers of
consumers' personal data. (5) The roles of controllers and processors
with respect to a consumer's personal data. (6) Requirements for data
protection assessments by controllers of consumers' personal data. (7)
Requirementsfor processing de-identified data or pseudonymous data.
(8) Limitations as to the scope of the new article. (9) The authority of
the attorney general to investigate and enforce suspected or actual
violations of the new article. (10) The preemption of local rules,
regulation, and laws regarding the processing of personal data.
Effective: January 1, 2025.
Brown L, Buchanan, Perfect, Koch,
Pol Jr.
January 12, 2022, read first time and referred to Committee onCommerce and Technology.
January 27, 2022, amended, reported favorably — Do Pass.
January 31, 2022, read second time, amended, ordered engrossed.
SB 358—LS 7186/DI 101

Reprinted
February 1, 2022
Second Regular Session of the 122nd General Assembly (2022)
PRINTING CODE. Amendments: Whenever an existing statute (or a section of the Indiana
Constitution) is being amended, the text of the existing provision will appear in this style type,
additions will appear in this style type, and deletions will appear in this style type.
Additions: Whenever a new statutory provision is being enacted (or a new constitutional
provision adopted), the text of the new provision will appear in this style type. Also, the
word NEW will appear in thatstyle type in the introductory clause of each SECTION that adds
a new provision to the Indiana Code or the Indiana Constitution.
Conflict reconciliation: Text in a statute in this style type or this style type reconciles conflicts
between statutes enacted by the 2021 Regular Session of the General Assembly.
SENATE BILL No. 358
A BILLFOR AN ACT to amend the Indiana Code concerning trade
regulation.
Be it enacted by the General Assembly of the State of Indiana:
1 SECTION 1. IC 24-15 IS ADDED TO THE INDIANA CODE AS
2 A NEW ARTICLE TO READ AS FOLLOWS [EFFECTIVE
3 JANUARY 1, 2025]:
4 ARTICLE 15. CONSUMER DATA PROTECTION
5 Chapter 1. Applicability
6 Sec. 1.(a) This article appliesto a person that conducts business
7 in Indiana or produces products or services that are targeted to
8 residents of Indiana and that:
9 (1) during a calendar year controls or processes personal data
10 of at least one hundred thousand (100,000) consumers; or
11 (2) controls or processes personal data of at least twenty-five
12 thousand (25,000) consumers and derives more than fifty
13 percent(50%) of gross revenue fromthe sale ofpersonaldata.
14 (b) This article does not apply to any:
15 (1) body, authority, board, bureau, commission, district, or
SB 358—LS 7186/DI 101
2
1 agency of the state or of any political subdivision of the state;
2 (2) financial institutions and affiliates, or data subject toTitle
3 V of the federal Gramm-Leach-Bliley Act (15 U.S.C. 6801 et
4 seq.);
5 (3) covered entity or business associate governed by the
6 privacy, security, and breach notification rules issued by the
7 United States Department of Health and Human Services(45
8 CFR Parts 160 and 164) pursuant to HIPAA;
9 (4) nonprofit organization;
10 (5) institution of higher education; or
11 (6) public utility (as defined in IC 8-1-2-1(a)) or service
12 company affiliated with a public utility (as defined in
13 IC 8-1-2-1(a)). For purposes of this subdivision, "service
14 company" means an associate company within a holding
15 company system organized specifically for the purpose of
16 providing goods or services to a public utility (as defined in
17 IC 8-1-2-1(a)) in the same holding company system.
18 Sec. 2.The following informationand data are exemptfromthis
19 article:
20 (1) Protected health information under HIPAA.
21 (2) Patient identifying information for purposes of 42 U.S.C.
22 290dd-2.
23 (3) Any of the following:
24 (A) Identifiable private information for purposes of the
25 federal policy for the protection of human subjects under
26 45 CFR Part 46.
27 (B) Identifiable private information that is otherwise
28 information collected as part of human subjects research
29 under the good clinical practice guidelines issued by the
30 International Council for Harmonisation of Technical
31 Requirements for Pharmaceuticals for Human Use.
32 (C) The protection of human subjects under 21 CFR Parts
33 6, 50, and 56.
34 (D) Personal data used or shared in research conducted in
35 accordance with the requirements set forth in this article.
36 (E) Other research conducted in accordance with
37 applicable law.
38 (4) Information and documents created for purposes of the
39 federal Health Care Quality Improvement Act of 1986 (42
40 U.S.C. 11101 et seq.).
41 (5) Patient safety work product for purposes of the federal
42 Patient Safety and Quality Improvement Act (42 U.S.C.
SB 358—LS 7186/DI 101
3
1 299b-21 et seq.).
2 (6) Information derived from any of the health care related
3 information set forth in this section that is de-identified in
4 accordance with the requirementsfor de-identificationunder
5 HIPAA.
6 (7) Information:
7 (A) originating from;
8 (B)intermingledwithso astobe indistinguishable from; or
9 (C) treated in the same manner as;
10 information that is exempt under this section and that is
11 maintained by a covered entity or business associate, as
12 defined in HIPAA, or a program or qualified service
13 organization, as defined in 42 U.S.C. 290dd-2.
14 (8) Information used only for public health activities and
15 purposes, as authorized by HIPAA.
16 (9) The collection, maintenance, disclosure, sale,
17 communication, or use of any personal information bearing
18 on a consumer's credit worthiness, credit standing, credit
19 capacity, character, general reputation, personal
20 characteristics, or mode of living by:
21 (A) a consumer reporting agency, furnisher, or user that
22 provides information for use in a consumer report; or
23 (B) a user of a consumer report;
24 but only to the extent that such activity is regulated by and
25 authorized under the federal Fair Credit Reporting Act (15
26 U.S.C. 1681 et seq.).
27 (10) Personal data collected, processed, sold, or disclosed in
28 compliance with the federal Driver's Privacy Protection Act
29 of 1994 (18 U.S.C. 2721 et seq.).
30 (11) Personal data regulated by the federal Family
31 Educational Rights and Privacy Act (20 U.S.C. 1232g etseq.).
32 (12) Personal data collected, processed, sold, or disclosed in
33 compliance with the federal Farm Credit Act (12 U.S.C. 2001
34 et seq.).
35 (13) Data processed or maintained:
36 (A)inthe course of anindividual applying to, employedby,
37 or acting as an agent or independent contractor of a
38 controller, processor, or third party, to the extent that the
39 data is collected and used within the context of that role;
40 (B) as emergency contact information for an individual
41 under this article and used for emergency contact
42 purposes; or
SB 358—LS 7186/DI 101
4
1 (C) that is necessary to retain to administer benefits for
2 another individual relating to the individual under clause
3 (A) and used for the purposes of administering those
4 benefits.
5 Sec. 3. A:
6 (1) controller; or
7 (2) processor;
8 that complies with the Children's Online Privacy Protection Act
9 (15 U.S.C. 6501 et seq.), and with any rules or regulations under
10 that act, satisfies any obligation to obtain parental consent under
11 this article.
12 Chapter 2. Definitions
13 Sec. 0.5. The definitions in this chapter apply throughout this
14 article.
15 Sec. 1. (a) "Affiliate" means a legal entity that:
16 (1) controls, is controlled by, or is under commoncontrolwith
17 another legal entity; or
18 (2) shares common branding with another legal entity.
19 (b) For purposes of this section, "control", with respect to a
20 company, means:
21 (1) ownership of, or the power to vote, more than fifty percent
22 (50%) ofthe outstanding shares of any class of voting security
23 of the company;
24 (2) control in any manner over the election of a majority of
25 the directors or ofindividuals exercising similar functions; or
26 (3) the power to exercise controlling influence over the
27 management of the company.
28 Sec. 2. "Aggregate data" means information:
29 (1) that relates to a group or category of consumers;
30 (2) from which individual consumer identities have been
31 removed; and
32 (3) that is not linked or reasonably linkable to any consumer.
33 Sec. 3. "Authenticate" means to verify through reasonable
34 meansthat a consumer who is entitledto exercise the personal data
35 rightsprovidedby IC24-15-3 isthe same consumer exercising such
36 rights with respect to particular personal data.
37 Sec. 4. (a) "Biometric data" means data that:
38 (1)is generatedby automaticmeasurements of anindividual's
39 biological characteristics, such as a fingerprint, a voiceprint,
40 images of the retina or iris, or other unique biological
41 patterns or characteristics; and
42 (2) is used to identify a specific individual.
SB 358—LS 7186/DI 101
5
1 (b) The term does not include:
2 (1) a physical or digital photograph;
3 (2) a video or audio recording, or data generated from a video
4 or audio recording; or
5 (3) information collected, used, or stored for health care
6 treatment, payment, or operations under HIPAA.
7 Sec. 5. "Business associate"hasthe meaning setforthin 45 CFR
8 160.103.
9 Sec. 6. "Child" means any individual who is less than thirteen
10 (13) years of age.
11 Sec. 7.(a) "Consent"means a clear affirmative actthatsignifies
12 a consumer's freely given, specific, informed, and unambiguous
13 agreement to process personal data relating to the consumer.
14 (b) For purposes of this section, a "clear affirmative act"
15 includes a written statement, including a statement written by
16 electronic means, or any other unambiguous affirmative action.
17 Sec. 8. (a) "Consumer" means an individual who:
18 (1) is a resident of Indiana; and
19 (2)is acting only for a personal,family, or householdpurpose.
20 (b) The term does not include an individual acting in a
21 commercial or employment context.
22 Sec. 9. "Controller" means a person that, alone or jointly with
23 others, determines the purpose and means of processing personal
24 data.
25 Sec. 10. "Covered entity" has the meaning set forth in 45 CFR
26 160.103.
27 Sec. 11. "Decision that produces legal or similarly significant
28 effects concerning a consumer" means a decision made by a
29 controller that results in the provision or denial by the controller
30 of:
31 (1) financial and lending services;
32 (2) housing;
33 (3) insurance;
34 (4) education enrollment;
35 (5) criminal justice;
36 (6) employment opportunities;
37 (7) health care services; or
38 (8) access to basic necessities, such as food and water.
39 Sec.12."De-identifieddata"meansdata that cannot reasonably
40 be linked to an identified or identifiable individual because a
41 controller that possesses the data:
42 (1) takes reasonable measures to ensure that the data cannot
SB 358—LS 7186/DI 101
6
1 be associated with an individual;
2 (2) publicly commits to maintaining and using the data
3 without attempting to re-identify the data; and
4 (3) obligates any recipients of the data through contractual
5 requirements to comply with all applicable provisions of this
6 article.
7 Sec. 13. "Health care provider" has the meaning set forth in
8 IC 4-6-14-2.
9 Sec. 14. "Health record" has the meaning set forth in
10 IC 1-1-4-5(a)(6).
11 Sec. 15. "HIPAA" refers to the federal Health Insurance
12 Portability and Accountability Act of 1996 (42 U.S.C. 1320d et
13 seq.).
14 Sec. 16. "Identified or identifiable individual" means an
15 individual who can be readily identified, directly or indirectly.
16 Sec. 17. "Institution of higher education" means a public or
17 private college or university.
18 Sec. 18. "Nonprofit organization" means any organization
19 exempt from taxation under Section 501(c)(3), 501(c)(6), or
20 501(c)(12) of the Internal Revenue Code.
21 Sec. 19.(a) "Personal data" meansinformation thatislinked or
22 reasonably linkable to an identified or identifiable consumer.
23 (b) The term does not include:
24 (1) de-identified data;
25 (2) aggregate data; or
26 (3) publicly available information.
27 Sec. 20. (a) "Precise geolocation data" means information
28 derived from technology, including global positioning system level
29 latitude and longitude coordinates, that directly identifies the
30 specific location of a natural person with precision and accuracy
31 within a radius of one thousand seven hundred fifty (1,750) feet.
32 (b) The term does not include the content of communications or
33 any data generated by or connected to advanced utility metering
34 infrastructure systems or equipment for use by a utility.
35 Sec. 21. "Processing", with respect to personal data, means any
36 operation or set of operations performed, whether by manual or
37 automated means, on personal data or on sets of personal data,
38 suchasthe collection, use,storage,disclosure, analysis, deletion, or
39 modification of personal data.
40 Sec. 22. "Processor" means a person that processes personal
41 data on behalf of a controller.
42 Sec. 23. "Profiling" means any form of solely automated
SB 358—LS 7186/DI 101
7
1 processing performed on personal data to evaluate, analyze, or
2 predict personal aspects related to an identified or identifiable
3 individual's economic situation, health, personal preferences,
4 interests, reliability, behavior, location, or movements.
5 Sec. 24. "Protected health information" has the meaning set
6 forth in 45 CFR 160.103.
7 Sec. 25."Pseudonymous data"meanspersonal data that cannot
8 be attributed to a specific individual because additional
9 information that would allow the data to be attributed to a specific
10 individual is:
11 (1) kept separately; and
12 (2) subject to appropriate technical and organizational
13 measures;
14 to ensure that the personal data is not attributed to an identified or
15 identifiable individual.
16 Sec. 26. "Publicly available information" means information:
17 (1) that is lawfully made available through federal, state, or
18 local government records; or
19 (2) that a business has a reasonable basisto believe islawfully
20 made available:
21 (A)to the generalpublic throughwidelydistributedmedia;
22 (B) by the consumer; or
23 (C) by a person to whom the consumer has disclosed the
24 information;
25 unless the consumer has restricted the information to a
26 specific audience.
27 Sec. 27. (a) "Sale of personal data" means the exchange of
28 personal data for monetary consideration by a controller to a third
29 party.
30 (b) The term does not include:
31 (1) the disclosure of personal data to a processor that
32 processes the personal data on behalf of the controller;
33 (2) the disclosure of personal data to a third party for
34 purposes of providing a product or service requested by the
35 consumer or the parent of a child;
36 (3) the disclosure or transfer of personal data to an affiliate of
37 the controller;
38 (4) the disclosure of information that the consumer:
39 (A) intentionally made available to the general public via
40 a channel of mass media; and
41 (B) did not restrict to a specific audience; or
42 (5) the disclosure or transfer of personal data to a third party
SB 358—LS 7186/DI 101
8
1 as an asset that is part of a proposed or actual merger,
2 acquisition, bankruptcy, or other transaction in which the
3 third party assumes control of all or part of the controller's
4 assets.
5 Sec. 28. "Sensitive data" means a category of personal data that
6 includes:
7 (1) personal data revealing racial or ethnic origin, religious
8 beliefs, a mental or physical health diagnosis, sexual
9 orientation, or citizenship or immigration status;
10 (2) genetic or biometric data that is processed for the purpose
11 of uniquely identifying a specific individual;
12 (3) personal data collected from a known child; and
13 (4) precise geolocation data.
14 Sec. 29. "State agency"hasthe meaning setforthin IC1-1-15-3.
15 Sec. 30. (a) "Targeted advertising" means the displaying of an
16 advertisementto a consumer inwhichthe advertisementisselected
17 based on personal data obtained from that consumer's activities
18 over time and across nonaffiliated websites or online applications
19 to predict the consumer's preferences or interests.
20 (b) The term does not include:
21 (1) advertisements based on activities within a controller's
22 own or affiliated websites or online applications;
23 (2) advertisements based on the context of a consumer's
24 currentsearch query, visit to a website, or online application;
25 (3) advertisements directed to a consumer in response to the
26 consumer's request for information or feedback; or
27 (4) the processing of personal data solely for measuring or
28 reporting advertising performance, reach, or frequency.
29 Sec. 31. "Third party", with respect to a context to which this
30 article applies, means a natural or legal person, public authority,
31 agency, or body other than:
32 (1) the consumer;
33 (2) the controller;
34 (3) the processor; or
35 (4) an affiliate of the processor or the controller.
36 Sec. 32. "Trade secret"hasthe meaning setforthin IC24-2-3-2.
37 Chapter 3. Personal Data; Consumer Rights
38 Sec. 1. (a) A consumer may invoke one (1) or more rights set
39 forth in subsection (b) by submitting to a controller a request
40 specifying the rights the consumer wishes to invoke. A known
41 child's parent or legal guardian may invoke on behalf of the child
42 one (1) or more rightsset forth in subsection (b) with respect to the
SB 358—LS 7186/DI 101
9
1 processing of personal data belonging to the known child by
2 submitting to a controller a request specifying the rights the
3 consumer wishes to invoke on behalf of the child. Except as
4 provided in IC 24-15-7-1(c) and IC 24-15-7-2, and subject to any
5 limitations or conditions set forth in subsections (b) and (c), a
6 controller shall comply with an authenticated consumer requestto
7 exercise a right set forth in subsection (b).
8 (b) A consumer has the following rights:
9 (1) To confirm whether or not a controller is processing the
10 consumer's personal data and, subject to the limitations set
11 forth in subdivision (4), to access such personal data.
12 (2) To correct inaccuracies in the consumer's personal data
13 that the consumer previously provided to a controller, taking
14 into account the nature of the personal data and the purposes
15 of the processing of the consumer's personal data. Upon
16 receiving a request from a consumer under this subdivision,
17 a controller shall correctinaccurate informationas requested
18 by the consumer, taking into account the nature of the
19 personal data and the purposes of the processing of the
20 consumer's personal data.
21 (3) To delete personal data provided by or obtained about the
22 consumer.
23 (4) To obtain either:
24 (A) a copy of; or
25 (B) a representative summary of;
26 the consumer's personal data that the consumer previously
27 provided to the controller. Information provided to a
28 consumer under thissubdivisionmust be in a portable and,to
29 the extent technically practicable, readily usable format that
30 allows the consumer to transmit the data or summary to
31 another controllerwithouthindrance,inany case inwhichthe
32 processing is carried out by automated means.The controller
33 has the discretion to send either a copy or a representative
34 summary of the consumer's personal data under this
35 subdivision, taking into account the nature of the personal
36 data and the purposes of the processing of the consumer's
37 personal data. A controller is not required to provide a copy
38 or a representative summary of a consumer's personal data
39 to the same consumer under this subdivision more than one
40 (1) time in a twelve (12) month period.
41 (5) To opt out of the processing of the consumer's personal
42 data for purposes of:
SB 358—LS 7186/DI 101
10
1 (A) targeted advertising;
2 (B) the sale of personal data; or
3 (C) profiling in furtherance of decisionsthat produce legal
4 or similarly significant effects concerning the consumer.
5 (c)Except as otherwiseprovidedinthis article, a controller shall
6 comply with a request by a consumer to exercise a consumer right
7 set forth in subsection (b) as follows:
8 (1) A controller shall respond to the consumer without undue
9 delay, but in any case not later than forty-five (45) days after
10 receipt of the consumer's request under this section. The
11 response period prescribed by this subdivision may be
12 extended once by an additional forty-five (45) days when
13 reasonably necessary,taking into accountthe complexity and
14 number of the consumer's requests, as long as the controller
15 informsthe consumer of any such extension within the initial
16 forty-five (45) day response period, along with the reason for
17 the extension.
18 (2) If a controller declines to take action regarding the
19 consumer's request, the controller shall inform the consumer
20 without undue delay, but in any case not later than forty-five
21 (45) days after receipt of the consumer's request under this
22 section, of the justification for declining to take action, and
23 shall provide instructions for how to appeal the decision
24 under subsection (d).
25 (3) Information provided in response to a consumer request
26 shall be provided by a controller free of charge, up to one (1)
27 time annually per consumer.If requestsfrom a consumer are
28 manifestly unfounded, excessive, or repetitive, the controller
29 may charge the consumer a reasonable fee to cover the
30 administrative costs of complying with the request or decline
31 to act on the request. The controller bears the burden of
32 demonstrating the manifestly unfounded, excessive, or
33 repetitive nature of the request.
34 (4) If a controller is unable to authenticate the request using
35 commercially reasonable efforts, the controller shall not be
36 required to comply with a request to initiate an action under
37 this section and may request that the consumer provide
38 additional information reasonably necessary to authenticate
39 the consumer and the consumer's request.
40 (d) A controller shall establish a process for a consumer to
41 appeal, within a reasonable period of time after the consumer's
42 receipt of a decision by the controller under subsection (c)(2), the
SB 358—LS 7186/DI 101
11
1 controller's refusal to take action on a request by the consumer
2 under this section. The appeal process shall be conspicuously
3 available and similar to the process for submitting requests to
4 invoke a right under this section. Not later than sixty (60) days
5 after receipt of an appeal, a controller shall inform the consumer
6 in writing of any action taken or not taken in response to the
7 appeal, including a written explanation of the reasons for the
8 decisions. If the appeal is denied, the controller shall also provide
9 the consumer with an online mechanism, if available, or other
10 method through which the consumer may contact the attorney
11 general to submit a complaint.
12 Chapter 4. Data Controller Responsibilities; Transparency
13 Sec. 1. Except as provided in IC 24-15-7-2, a controller has the
14 following responsibilities:
15 (1) A controller shall limit the collection of personal data to
16 what is adequate, relevant, and reasonably necessary in
17 relation to the purposes for which such data is processed, as
18 disclosed to the consumer.
19 (2) Except as otherwise provided in this article, a controller
20 shall not process personal data for purposes that are neither
21 reasonably necessary to nor compatible with the disclosed
22 purposes for which the personal data is processed, unless the
23 controller obtains the consumer's consent.
24 (3) A controller shall establish, implement, and maintain
25 reasonable administrative, technical, and physical data
26 security practicesto protectthe confidentiality,integrity, and
27 accessibility of personal data. The data security practices
28 required under this subdivision must be appropriate to the
29 volume and nature of the personal data at issue.
30 (4) A controller shall not process personal data in violation of
31 state and federal laws that prohibit unlawful discrimination
32 against consumers.Acontroller shallnotdiscriminate against
33 a consumer for exercising any ofthe consumer rightssetforth
34 in this article, including by denying goods or services to the
35 consumer, charging different prices or rates for goods and
36 services, or providing a different level or quality of goods or
37 servicestothe consumer.However,nothing inthissubdivision
38 shall be construed to:
39 (A) require a controller to provide a product or service
40 that requires the personal data of a consumer that the
41 controller does not collect or maintain; or
42 (B) prohibit a controller from offering a different price,
SB 358—LS 7186/DI 101
12
1 rate, level, quality, or selection of goods or services to a
2 consumer, including offering goods or services for no fee,
3 if the consumer has exercised the consumer's right to opt
4 out under IC 24-15-3-1(b)(5) or if the offer is related to a
5 consumer's voluntary participation in a bona fide loyalty,
6 rewards, premium features, discount, or club card
7 program.
8 (5) A controller shall not process sensitive data concerning a
9 consumerwithout obtaining the consumer's consent, or,inthe
10 case of the processing of sensitive data concerning a known
11 child, without processing such data in accordance with the
12 federal Children's Online Privacy Protection Act (15 U.S.C.
13 6501 et seq.).
14 Sec. 2. Any provision of a contract or agreement of any kind
15 that purports to waive or limit in any way a consumer's rights
16 under IC 24-15-3 is contrary to public policy and is void and
17 unenforceable.
18 Sec. 3. A controller shall provide consumers with a reasonably
19 accessible, clear, and meaningful privacy notice that includes:
20 (1)the categories ofpersonaldataprocessedbythe controller;
21 (2) the purpose for processing personal data;
22 (3) how consumers may exercise their consumer rights under
23 IC 24-15-3, including how a consumer may appeal a
24 controller's decision with regard to the consumer's request;
25 (4) the categories of personal data that the controller shares
26 with third parties, if any; and
27 (5) the categories of third parties, if any, with whom the
28 controller shares personal data.
29 Sec. 4. If a controller sells a consumer's personal data to third
30 parties oruses a consumer'spersonaldata for targetedadvertising,
31 the controller shall clearly andconspicuouslydisclose suchactivity,
32 as well as the manner in which a consumer may exercise the right
33 to opt out of such sales or use.
34 Sec. 5. A controller shall establish, and shall describe in a
35 privacy notice provided under section 3 of this chapter, one (1) or
36 more secure and reliable meansfor consumersto submit a request
37 to exercise their rights under IC 24-15-3. Such means must take
38 into account:
39 (1) the ways in which consumers normally interact with the
40 controller;
41 (2)the needfor the secure andreliable communicationofsuch
42 requests; and
SB 358—LS 7186/DI 101
13
1 (3) the ability of the controller to authenticate the identity of
2 the consumer making the request.
3 A controller may not require a consumer to create a new account
4 in order to exercise the consumer's rights under IC 24-15-3 but
5 may require a consumer to use an existing account.
6 Chapter 5. Responsibility According to Role; Controllers and
7 Processors
8 Sec. 1. A processor shall adhere to the instructions of a
9 controller and shall assist the controller in meeting its obligations
10 under this chapter. Such assistance shall include the following:
11 (1) Fulfilling the controller's obligation to respond to
12 consumer requestsunder IC24-15-3by appropriate technical
13 and organizational measures, insofar as this is reasonably
14 practicable, and taking into account the nature of processing
15 and the information available to the processor.
16 (2) Taking into account the nature of processing and the
17 informationavailable to theprocessor,assisting the controller
18 in meeting the controller's obligations in relation to:
19 (A) the security of processing the personal data; and
20 (B) the notification of a breach of security of the system of
21 the processor under IC 24-4.9;
22 in order to meet the controller's obligations.
23 (3) Providing necessary information to enable the controller
24 to conduct and document data protection assessments under
25 IC 24-15-6.
26 Sec. 2. (a) A contract between a controller and a processor shall
27 govern the processor's data processing procedures with respect to
28 processing performed on behalf of the controller. The contract
29 must be binding and clearly set forth instructions for processing
30 personal data, the nature and purpose of processing, the type of
31 data subject to processing, the duration of processing, and the
32 rights and obligations of both parties. The contract must also
33 include requirements that the processor do the following:
34 (1) Ensure that each individual processing personal data is
35 subject to a duty of confidentiality with respect to the data.
36 (2) At the controller's direction, delete or return all personal
37 data to the controller as requested at the end of the provision
38 ofservices,unless retentionofthepersonaldata is requiredby
39 law.
40 (3) Upon the reasonable request of the controller, make
41 available to the controller all information in its possession
42 necessary to demonstrate the processor's compliance withthe
SB 358—LS 7186/DI 101
14
1 obligations in this chapter.
2 (4) Allow, and cooperate with, reasonable assessments by the
3 controller or the controller's designated assessor.
4 Alternatively, the processor may arrange for a qualified and
5 independent assessor to conduct an assessment of the
6 processor's policies and technical and organizational
7 measures in support of the processor's obligations under this
8 chapter using an appropriate and accepted control standard
9 or framework and assessment procedure for such
10 assessments.The processor shall provide a report of any such
11 assessment to the controller upon request.
12 (5) Subject to subsection (b), engage any subcontractor
13 pursuantto awrittencontractthat requiresthe subcontractor
14 to meet the obligations of the processor with respect to the
15 personal data.
16 (b) Nothing in this section shall be construed to relieve a
17 controller or a processor from the liabilities imposed on the
18 controller or processor by virtue of its role in the processing
19 relationship.
20 Sec. 3. Determining whether a person is acting as a controller or
21 a processor with respect to a specific processing of data is a fact
22 based determination that depends upon the context in which
23 personal data is processed. A processor that continuesto adhere to
24 a controller's instructions with respect to a specific processing of
25 personal data remains a processor.
26 Chapter 6. Data Protection Assessments
27 Sec. 1. (a) The data protection assessment requirements set
28 forth in this chapter apply to processing activities created or
29 generated after December 31, 2024, and are not retroactive to any
30 processing activities created or generated before January 1, 2025.
31 (b) A controller shall conduct and document a data protection
32 assessment of each of the following processing activities involving
33 personal data:
34 (1) The processing of personal data for purposes of targeted
35 advertising.
36 (2) The sale of personal data.
37 (3) The processing of personal data for purposes of profiling,
38 if such profiling presents a reasonably foreseeable risk of:
39 (A) unfair or deceptive treatment of, or unlawful disparate
40 impact on, consumers;
41 (B) financial, physical, or reputational injury to
42 consumers;
SB 358—LS 7186/DI 101
15
1 (C) a physical or other intrusion upon the solitude or
2 seclusion, or the private affairs or concerns, of consumers,
3 ifsuchintrusionwouldbe offensive to a reasonableperson;
4 or
5 (D) other substantial injury to consumers.
6 (4) The processing of sensitive data.
7 (5) Any processing activities involving personal data that
8 present a heightened risk of harm to consumers.
9 (c) Data protection assessments conducted under this chapter
10 shall identify and weigh the benefits that may flow, directly and
11 indirectly, from the processing to the controller, the consumer,
12 other stakeholders, and the public against the potential risksto the
13 rights of the consumer associated with such processing, as
14 mitigated by safeguards that can be employed by the controller to
15 reduce such risks.The use of de-identified data and the reasonable
16 expectations of consumers, as well as the context of the processing
17 and the relationship between the controller and the consumer
18 whose personal data will be processed, shall be factored into this
19 assessment by the controller.
20 (d) A single data protection assessment may address a
21 comparable set of processing operations that include similar
22 activities.
23 (e) A data protection assessment conducted by a controller for
24 the purpose of compliance with other laws or regulations may be
25 used to comply with thissection if the assessment has a reasonably
26 comparable scope andeffectto anassessment conductedunder this
27 section.
28 Sec. 2. (a) The attorney general may request, pursuant to a civil
29 investigativedemand,that a controllerdisclose anydataprotection
30 assessment that is relevant to an investigation conducted by the
31 attorney general. Upon receipt of such a request, the controller
32 shall make the data protectionassessment available to the attorney
33 general. Subject to subsection (b), the attorney general may
34 evaluate the data protection assessment for a controller's
35 compliance with the responsibilities set forth in IC 24-15-4.
36 (b) Data protection assessments are confidential and exempt
37 from public inspection and copying under IC 5-14-3-4. The
38 disclosure of a data protection assessment pursuant to a request
39 from the attorney general does not constitute a waiver of
40 attorney-client privilege or work product protection with respect
41 to the assessment andany informationcontainedinthe assessment.
42 Chapter 7. Processing De-identified Data or Pseudonymous
SB 358—LS 7186/DI 101
16
1 Data; Exemptions
2 Sec. 1. (a) A controller in possession of de-identified data shall:
3 (1) take reasonable measures to ensure that the data cannot
4 be associated with an individual;
5 (2) publicly commit to maintaining and using de-identified
6 data without attempting to re-identify the data; and
7 (3) contractually obligate any recipients of the de-identified
8 data to comply with all provisions of this chapter.
9 (b) This chapter shall not be construed to require a controller
10 or processor to:
11 (1) re-identify de-identified data or pseudonymous data;
12 (2) maintain data in identifiable form; or
13 (3) collect, obtain, retain, or access any data or technology, in
14 order to be capable of associating an authenticated consumer
15 request with personal data.
16 (c)This chapter shall not be construed to require a controller or
17 processor to complywitha request of a consumer under IC24-15-3
18 if all of the following conditions are met:
19 (1) The controller is not reasonably capable of associating the
20 request with the personal data or it would be unreasonably
21 burdensome for the controller to associate the request with
22 the personal data.
23 (2) The controller does not use the personal data to recognize
24 or respond to the specific consumer who is the subject of the
25 personal data, or associate the personal data with other
26 personal data about the same specific consumer.
27 (3) The controller does not sell the personal data to any third
28 party or otherwise voluntarily disclose the personal data to
29 any third party other than a processor.
30 Sec. 2. The:
31 (1) rights of a consumer set forth in IC 24-15-3-1(b)(1)
32 through IC 24-15-3-1(b)(4); and
33 (2) responsibilities of a controller under IC 24-15-4-1(1)
34 through IC 24-15-4-1(4);
35 do not apply to pseudonymous data in any case in which the
36 controller is able to demonstrate that any informationnecessary to
37 identify the consumer is kept separately and is subject to effective
38 technical and organizational controls that prevent the controller
39 from accessing such information.
40 Sec. 3. A controller that discloses pseudonymous data or
41 de-identified data shall exercise reasonable oversight to monitor
42 compliance with any contractual commitments to which the
SB 358—LS 7186/DI 101
17
1 pseudonymous data or de-identified data is subject and shall take
2 appropriate steps to address any breaches of those contractual
3 commitments.
4 Chapter 8. Limitations
5 Sec. 1. (a) This article shall not be construed to restrict a
6 controller's or processor's ability to do any of the following:
7 (1) Comply with federal, state, or local laws, rules, or
8 regulations.
9 (2) Comply with a civil, criminal, or regulatory inquiry,
10 investigation,subpoena, or summons by a federal,state, local,
11 or other governmental authority.
12 (3) Cooperate with law enforcement agencies concerning
13 conduct or activity thatthe controllerorprocessor reasonably
14 and in good faith believes may violate federal, state, or local
15 laws, rules, or regulations.
16 (4)Investigate, establish, exercise,prepare for, ordefendlegal
17 claims.
18 (5) Provide a product or service specifically requested by a
19 consumer, perform a contract to which the consumer, or a
20 parent of a child, is a party, including fulfilling the terms of a
21 writtenwarranty, or take steps atthe request ofthe consumer
22 or parent before entering into a contract.
23 (6) Take immediate steps to protect an interest that is
24 essential for the life or physical safety of the consumer or of
25 another individual, if the processing cannot be manifestly
26 based on another legal basis.
27 (7) Prevent, detect, protect against, or respond to security
28 incidents, identity theft, fraud, harassment, malicious or
29 deceptive activities, or any illegal activity, to preserve the
30 integrity or security of systems, or to investigate, report, or
31 prosecute those responsible for any such action.
32 (8) Engage in public or peer reviewed scientific or statistical
33 research that is in the public interest and that adheres to all
34 applicable ethics and privacy laws and is approved,
35 monitored, and governed by an institutional review board, or
36 a similar independent oversight entity, that determines if:
37 (A)the informationislikely to provide substantial benefits
38 that do not exclusively accrue to the controller;
39 (B) the expected benefits of the research outweigh the
40 privacy risks; and
41 (C) the controller hasimplemented reasonable safeguards
42 to mitigate privacy risks associated with research,
SB 358—LS 7186/DI 101
18
1 including any risks associated with re-identification.
2 (9) Assist another controller, processor, or third party with
3 any obligation described in this section.
4 (b) Processing personal data for a purposes expressly identified
5 in subsection (a)(1) through (a)(9) does not by itself make a person
6 a controller with respect to such processing.
7 Sec. 2. The obligations imposed on a controller or a processor
8 under this article do not restrict a controller's or processor's
9 ability to collect, use, or retain data to do the following:
10 (1) Conduct internal research to develop, improve, or repair
11 products, services, or technology.
12 (2) Effectuate a product recall.
13 (3)Identify andrepair technical errorsthatimpair existing or
14 intended functionality.
15 (4) Perform internal operations that are:
16 (A) reasonably aligned with the expectations of the
17 consumer;
18 (B) reasonably anticipated based on the consumer's
19 existing relationship with the controller; or
20 (C) otherwise compatible with:
21 (i) processing data in furtherance of the provision of a
22 product or service specifically requested by a consumer,
23 or the parent of a child; or
24 (ii) the performance of a contract to which the consumer
25 is a party.
26 Sec. 3. The obligations imposed on a controller or a processor
27 under this article do not apply if compliance by the controller or
28 processor with this article would violate an evidentiary privilege
29 under Indiana law. This article shall not be construed to prevent
30 a controller or processor from providing, as part of a privileged
31 communication, personal data concerning a consumer to a person
32 covered by an evidentiary privilege under Indiana law.
33 Sec. 4. A controller or processor that discloses personal data to
34 a thirdparty controller or processor incompliancewiththis article
35 is not in violation of this article if the third party controller or
36 processor that receives and processes the personal data violates
37 this article, as long as, at the time of disclosing the personal data,
38 the disclosing controller or processor did not have actual
39 knowledge that the recipient intended to commit a violation. A
40 third party controller or processor receiving personal data from a
41 controller or processor is likewise not in violation of this article
42 solely because of the transgressions of the controller or processor
SB 358—LS 7186/DI 101
19
1 from which it receives such personal data.
2 Sec. 5. This article:
3 (1) shall not be construed as an obligation imposed on
4 controllers and processorsthat adversely affectsthe rights or
5 freedoms of any persons, such as exercising the right of free
6 speech under the First Amendment to the Constitution of the
7 United States; and
8 (2) does not apply to personal data in the context of a purely
9 personal or household activity.
10 Sec. 6. Nothing in this article shall be construed as requiring a
11 controller to disclose trade secrets.
12 Sec. 7. (a) Personal data processed by a controller for a purpose
13 authorized under this chapter may not be processed for any other
14 purpose unless otherwise allowed under this article. Personal data
15 processed by a controller under this chapter may be processed to
16 the extent that such processing is:
17 (1) reasonably necessary and proportionate to a purpose
18 authorized under this chapter; and
19 (2) adequate, relevant, and limited to what is necessary in
20 relation to the specific purpose.
21 (b) Personal data collected, used, or retained under section 2 of
22 this chapter shall, as applicable, take into account the nature and
23 purpose of the collection, use, or retention. Any personal data
24 collected, used, or retained must be subject to reasonable
25 administrative, technical, and physical measures to:
26 (1) protect the confidentiality, integrity, and accessibility of
27 the personal data; and
28 (2) reduce reasonably foreseeable risks of harm to consumers
29 relating to such collection, use, or retention of the personal
30 data.
31 (c) If a controller processes personal data pursuant to an
32 exemption under this chapter, the controller bears the burden of
33 demonstrating that such processing:
34 (1) qualifies for the exemption; and
35 (2) complies with the requirements set forth in this section.
36 Chapter 9. Investigative Authority
37 Sec. 1. Whenever the attorney general has reasonable cause to
38 believe that any person has engaged in, is engaging in, or is about
39 to engage in any violation of this article, the attorney general is
40 empowered to issue a civil investigative demand to investigate the
41 suspected violation.
42 Chapter 10. Enforcement
SB 358—LS 7186/DI 101
20
1 Sec. 1. The attorney general has exclusive authority to enforce
2 the provisions of this article.
3 Sec. 2. (a) Before initiating an action under this chapter, the
4 attorney generalshall provide a controller or processor thirty (30)
5 dayswrittennotice identifying the specific provisions ofthis article
6 that the attorney general alleges have been or are being violated.
7 If within the thirty (30) day period set forth in this section, the
8 controller or processor:
9 (1) cures the alleged violation; and
10 (2)providestheattorney general anexpresswrittenstatement
11 that:
12 (A) the alleged violation has been cured; and
13 (B) actions have been taken to ensure no further such
14 violations will occur;
15 the attorney general shall not initiate an action against the
16 controller or processor.
17 (b) If a controller or processor:
18 (1) continues to violate this chapter following the thirty (30)
19 day period set forth in subsection (a); or
20 (2) breaches an express written statement provided to the
21 attorney general under subsection (a)(2);
22 the attorney general may initiate an action in the name of the state
23 and may seekaninjunction to restrain any violations of this article
24 and a civil penalty not to exceed seven thousand five hundred
25 dollars ($7,500) for each violation under this article.
26 (c) The attorney general may recover reasonable expenses
27 incurred in investigating and preparing the case, including
28 attorney fees, in any action initiated under this chapter.
29 Sec. 3. Nothing inthis article shall be construed as providing the
30 basis for a private right of action for violations of this article or
31 any other law.
32 Chapter 11. Preemption; Other Laws
33 Sec. 1. This article supersedes and preempts all rules,
34 regulations, codes, ordinances, and other laws adopted by a city,
35 county, city and county, municipality, or local agency regarding
36 the processing of personal data by controllers or processors.
37 Sec. 2. Any reference to federal, state, or local law or statute in
38 this article includes any accompanying rules, regulations, or
39 exemptions.
SB 358—LS 7186/DI 101
21
COMMITTEE REPORT
Madam President: The Senate Committee on Commerce and
Technology, to which was referred Senate Bill No. 358, has had the
same under consideration and begsleave to report the same back to the
Senate with the recommendation that said bill be AMENDED as
follows:
Delete everything afterthe enacting clause and insert the following:
(SEE TEXT OF BILL)
and when so amended that said bill do pass.
(Reference is to SB 358 as introduced.)
PERFECT, Chairperson
Committee Vote: Yeas 10, Nays 0.
_____
SENATE MOTION
Madam President: I move that Senate Bill 358 be amended to read
as follows:
Replace the effective date in SECTION 1 with "[EFFECTIVE
JANUARY 1, 2025]".
Page 2, line 8, delete "HIPPA;" and insert "HIPAA;".
Page 2, line 9, delete "or".
Page 2, line 10, delete "education." and insert "education; or
(6) public utility (as defined in IC 8-1-2-1(a)) or service
company affiliated with a public utility (as defined in
IC 8-1-2-1(a)). For purposes of this subdivision, "service
company" means an associate company within a holding
company system organized specifically for the purpose of
providing goods or services to a public utility (as defined in
IC 8-1-2-1(a)) in the same holding company system.".
Page 2, line 13, delete "HIPPA." and insert "HIPAA.".
Page 2, line 40, delete "HIPPA." and insert "HIPAA.".
Page 3, line 5, delete "HIPPA," and insert "HIPAA,".
Page 3, line 8, delete "HIPPA." and insert "HIPAA.".
Page 5, delete line 42.
Page 6, delete line 1.
Page 6, line 2, delete "14." and insert "13.".
Page 6, line 4, delete "15." and insert "14.".
SB 358—LS 7186/DI 101
22
Page 6, line 6, delete "16." and insert "15.".
Page 6, line 9, delete "17." and insert "16.".
Page 6, line 11, delete "18." and insert "17.".
Page 6, line 13, delete "19." and insert "18.".
Page 6, line 16, delete "20." and insert "19.".
Page 6, line 22, delete "21." and insert "20.".
Page 6, line 27, delete "in".
Page 6, line 28, delete "connection with,".
Page 6, line 28, delete "to," and insert "to".
Page 6, line 31, delete "22." and insert "21.".
Page 6, line 36, delete "23." and insert "22.".
Page 6, line 38, delete "24." and insert "23.".
Page 7, line 1, delete "25." and insert "24.".
Page 7, line 3, delete "26." and insert "25.".
Page 7, line 12, delete "27." and insert "26.".
Page 7, line 23, delete "28." and insert "27.".
Page 8, line 1, delete "29." and insert "28.".
Page 8, line 10, delete "30." and insert "29.".
Page 8, line 11, delete "31." and insert "30.".
Page 8, line 25, delete "32." and insert "31.".
Page 8, line 32, delete "33." and insert "32.".
Page 8, line 36, after "A" insert "known".
Page 8, line 39, after "the" insert "known".
Page 12, line 6, after "a" insert "known".
Page 12, line 11, delete "consumer right's" and insert "consumer's
rights".
Page 14, line 25, delete "2023," and insert "2024,".
Page 14, line 26, delete "2024." and insert "2025.".
Page 15, line 33, delete "IC 5-14-13-4." and insert "IC 5-14-3-4.".
Page 19, line 2, delete "United States" and insert "Constitution of
the United States;".
Page 19, line 3, delete "Constitution;"
Page 20, line 9, after "further" insert "such".
Page 20, delete lines 28 through 42.
Page 21, line 1, delete "12." and insert "11.".
(Reference is to SB 358 as printed January 28, 2022.)
BROWN L
SB 358—LS 7186/DI 101