LRB-6076/1
KP&MED:cdc
2021 - 2022 LEGISLATURE
2021 SENATE BILL 997
February 17, 2022 - Introduced by LAW REVISION COMMITTEE. Referred to
Committee on Senate Organization.
AN ACT to amend 14.87 (title), 227.21 (2) (b), 441.06 (4), 441.10 (7), subchapter
II (title) of chapter 441 [precedes 441.51] and 441.51 (title) of the statutes;
relating to: incorporation of standards by reference into agency
administrative rules and the nurse licensure compact (suggested as remedial
legislation by the Department of Safety and Professional Services).
Analysis by the Legislative Reference Bureau
Current law generally requires all permanent rules that are promulgated by
state agencies to be published in the Wisconsin Administrative Code. However,
current law allows an agency, in order to avoid unnecessary expense, to adopt
standards established by technical societies and organizations of recognized
national standing by incorporating the standards in its rules by reference to the
specific issue or issues of the publication in which they appear, without reproducing
the standards in full in their rules. Such incorporation by reference requires the
consent of the attorney general and requires the rule incorporating the standard to
state that the standards are on file at the offices of the agency and the Legislative
Reference Bureau.
This bill deletes the requirement that the agency state that the standards are
on file at the offices of the agency and instead requires only that the agency state that
the standards are on file at the Legislative Reference Bureau.
The bill also makes terminology changes related to the Nurse Licensure
Compact. Specifically, the bill removes the word “enhanced” from references in the
1
2
3
4
5
2021 - 2022 Legislature - 2 - LRB-6076/1
KP&MED:cdc
 SENATE BILL 997
statutes to the Nurse Licensure Compact. Under current law, Wisconsin has entered
into the Nurse Licensure Compact, which allows a registered nurse or licensed
practical nurse who is licensed in a state that has adopted the compact to practice
in any other state that has adopted the compact without needing to separately obtain
a license in that other state.
For further information, see the NOTES provided by the Law Revision
Committee of the Joint Legislative Council.
The people of the state of Wisconsin, represented in senate and assembly, do
enact as follows:
LAW REVISION COMMITTEE PREFATORY NOTE: This bill is a remedial legislation
proposal, requested by the Department of Safety and Professional Services and
introduced by the Law Revision Committee under s. 13.83 (1) (c) 4. and 5., stats. After
careful consideration of the various provisions of the bill, the Law Revision Committee
has determined that this bill makes minor substantive changes in the statutes, and that
these changes are desirable as a matter of public policy.
SECTION 1. 14.87 (title) of the statutes is amended to read:
14.87 (title) Enhanced nurse Nurse licensure compact.
SECTION 2. 227.21 (2) (b) of the statutes is amended to read:
227.21 (2) (b) The attorney general shall consent to incorporation by reference
only in a rule of limited public interest and in a case where the incorporated
standards are readily available in published form or are available on optical disc or
in another electronic format. Each rule containing an incorporation by reference
shall state how the material incorporated may be obtained and, except as provided
in s. 601.41 (3) (b), that the standards are on file at the offices of the agency and the
legislative reference bureau.
NOTE: This SECTION deletes a requirement that an agency maintain, at its offices,
a copy of any standards established by technical societies and national organizations that
are incorporated into its rules but not reproduced in full.
SECTION 3. 441.06 (4) of the statutes is amended to read:
441.06 (4) Except as provided in s. 257.03, no person may practice or attempt
to practice professional nursing, nor use the title, letters, or anything else to indicate
that he or she is a registered or professional nurse unless he or she is licensed under
1
2
3
4
5
6
7
8
9
10
11
12
13
14
2021 - 2022 Legislature - 3 - LRB-6076/1
KP&MED:cdc
SECTION 3 SENATE BILL 997
this section. Except as provided in s. 257.03, no person not so licensed may use in
connection with his or her nursing employment or vocation any title or anything else
to indicate that he or she is a trained, certified or graduate nurse. This subsection
does not apply to any registered nurse who holds a multistate license, as defined in
s. 441.51 (2) (h), issued by a jurisdiction, other than this state, that has adopted the
enhanced nurse licensure compact under s. 441.51.
NOTE: SECTIONS 1 and 3 to 6 update references to the national nursing compact to
eliminate the term “enhanced” from the title.
SECTION 4. 441.10 (7) of the statutes is amended to read:
441.10 (7) No license is required for practical nursing, but, except as provided
in s. 257.03, no person without a license may hold himself or herself out as a licensed
practical nurse or licensed attendant, use the title or letters “Trained Practical
Nurse" or “T.P.N.", “Licensed Practical Nurse" or “L.P.N.", “Licensed Attendant" or
“L.A.", “Trained Attendant" or “T.A.", or otherwise seek to indicate that he or she is
a licensed practical nurse or licensed attendant. No licensed practical nurse or
licensed attendant may use the title, or otherwise seek to act as a registered, licensed,
graduate or professional nurse. Anyone violating this subsection shall be subject to
the penalties prescribed by s. 441.13. The board shall grant without examination a
license as a licensed practical nurse to any person who was on July 1, 1949, a licensed
attendant. This subsection does not apply to any licensed practical nurse who holds
a multistate license, as defined in s. 441.51 (2) (h), issued by a jurisdiction, other than
this state, that has adopted the enhanced nurse licensure compact under s. 441.51.
SECTION 5. Subchapter II (title) of chapter 441 [precedes 441.51] of the statutes
is amended to read:
CHAPTER 441
1
2
3
4
5
6
7
8
9
10
11
12
13
14
15
16
17
18
19
20
21
22
23
2021 - 2022 Legislature - 4 - LRB-6076/1
KP&MED:cdc
 SENATE BILL 997 SECTION 5
SUBCHAPTER II
ENHANCED NURSE LICENSURE COMPACT
SECTION 6. 441.51 (title) of the statutes is amended to read:
441.51 (title) Enhanced nurse Nurse licensure compact.
(END)
1
2
3
4
5